Keep your Laziness in Check

### 2018-03-08 paper/main.tex
[paper] more on section 2 and 3

We can write another executable specification for this property of
$\lk{solutionStep}_n$. To describe the concrete \textit{demand} on
maps, we need to first give a data type definition for \lk{Map}. We
choose a simple binary search tree representation here. We also define
a \lk{lookupD} function \hzh{we should probably move this to the
appendix} to lookup demand at a given key.
%
\begin{inlinecode}
data Map  k v = Bin (Map k v) k v (Map k v)
              | Empty

data MapD k v = BinD (Thunk (MapD k v)) (Thunk k) (Thunk v) (Thunk (MapD k v))
              | EmptyD

lookupD :: k -> MapD k v -> Thunk v
lookupD k EmptyD                          = T
lookupD k (BinD _ T      _ _)             = T
lookupD k (BinD l (E k') v r) | k == k'   = v
                              | k <= k'   = recur l
                              | otherwise = recur r
  where recur (E demand) = lookupD k demand
        recur T          = T
\end{inlinecode}
%
Now, we can write an executable specification
for \lk{solutionStep}. \hzh{TODO: write the spec with
strictCheckWithResults, and show a counterexample}
%
\begin{inlinecode}
\end{inlinecode}
%

\leo{Punned names for sections? Or is that just an urns thing? :P }

\begin{itemize}
\item In this section we will showcase our proposed specification language \leo{It's not a language tho}
\item while introducing the running example of our paper: maps.
\item Maps are particularly well-suited for a discussion of laziness because of the interesting different
possible behaviors: strict-maps, value-lazy maps \leo{Point out Data.Map.Strict
vs Data.Map}, and fully lazy maps
\item For simplicity, we will use binary search trees as maps, assuming the keys are integers and forego balancing
as the added complexity is irrelevant to the laziness discussion. \leo{Can we claim it wouldn't matter at all? I think so}
\end{itemize}

\section{Demand Types and Specification}
\label{sec:demand-types-and-spec}
\hzh{In this section, we should explain how demand types and specification work in general. We don't actually have to define a ListD/TreeD for every type out there. There is also a generic pattern for the type of a specification given the type of a function. We should also talk about the non-exact versions of specifications, and how they work}

There is a general pattern between a type \lk{a} and
its \textit{demand type}, for which we will write \lk{Demand a} from
now on. In both the example on lists and the example on maps,
a \lk{Thunk} type constructor is placed around all of the constructor
arguments to the original type. The \lk{Thunk} type has the following
definition:
\begin{inlinecode}
data Thunk a = E !a | T
\end{inlinecode}
Its two constructors \lk{E} and \lk{T} represents an evaluated
fragment of the original data, and unexplored regions of the original
data, respectively.
%
This design is motivated by the runtime representation of Haskell
values. Boxed values in Haskell are indirectly represented by a
pointer, and the pointer can lead to either a constructor bundled with
pointers to deeper parts of the value, or a piece of unevaluated code,
which will produce a fragment of the value upon execution. This piece
of code is also know as a thunk.

@@

A \textit{specification} in StrictCheck is a function that computes
demand values. When we talk about Haskell function's laziness
informally, we often refer to these functions as ``strict'' or
``lazy'' in its arguments. There is a lot of information we can unpack
from such phrases --- how strict the function is in its arguments, the
dependency of the strictness on the argument values, and the
relationship between demand on inputs with demand on the output.

@@

In general, specification types are not unique for the same
function. Again consider \lk{take}, we can certainly choose the
specification type \lk{Spec '[Int] ([a] -> [a])} as well. However, the
meaning of this specification is very different from the other
choice. A specification of this type predicts how the \lk{Int}
argument is used given a demand on the partially-applied \lk{take}
function, and this specification does not produce information on how
the lists are used at all.

StrictCheck specifications apply to higher-order functions, and are
compositional. Take the example of \lk{map :: (a -> b) -> [a] -> [b]},
the type of its specification is \lk{Spec '[a] b -> Spec '[[a]] [b]},
a function from specifications to specifications. This allows
programmers to apply the specification of \lk{map} to any
specification of the function argument to \lk{map}, and conveniently
synthesize a specification based on that of \lk{map}.

The \lk{Spec} type also has a single constructor with the same
name. This constructor takes a higher-order function in
continuation-passing-style ~\cn, where the first argument to the
function is a framework-supplied function that takes the predicted
demand on the inputs. We will always name this argument \lk{predict}
in this paper. The function is CPS'd to hide some internal plumbing
that presents a variadic interface to the user, so that the number of
arguments to \lk{predict} always matches the length of the type-level
list in \lk{Spec}.

@@

>>>>

We can write another executable specification for this property of
$\lk{solutionStep}_n$. To describe the concrete \textit{demand} on
maps, we need to first give a data type definition for \lk{Map}. We
choose a simple binary search tree representation here. We also define
a function \lk{iterSolution} that performs $n$ iterations
of \lk{solutionStep}, and indexes into the map at a given key. The
specification is written for \lk{iterSolution}.
%
\begin{inlinecode}
data Map  k v = Bin (Map k v) k v (Map k v)
              | Empty

iterSolution :: Key -> Int -> Map (Int, Int) Int -> Maybe Int
iterSolution (Key k) n soln = lookup m k
  where m | n <= 0    = soln
          | otherwise = (iterate solutionStep soln) !! n

iterSolution_spec :: Evaluation '[Key, Int, KMap] (Maybe Int)
                  -> Maybe (Int, Int)
iterSolution_spec (Evaluation args demands dOut) =
  let I (Key evalK) = hd args
      I nIter       = hd (tl args)
      dInM          = hd (tl (tl demands))
      inM           = replaceThunk Empty (fromDemand @KMap dInM)
  in  if (inM == Empty)   ||
         isBaseCase evalK ||
         nIter <= 0       ||
         (isNothing $ lookup inM evalK)
      then Nothing
      else Just evalK
  where isBaseCase (j, k) = j - 1 < 0 || k - weights !! j < 0
\end{inlinecode}
%
This specification is different from \lk{rot\_spec} --- it uses
the \lk{Evaluation} type instead of \lk{Spec}. In StrictCheck, we
reverse the \lk{Spec} type for specifications that computes the exact
demand on inputs. However, there many cases where we might want to
only check whether some property holds of the observed demands on the
inputs. For example, here we only need to check that the value
at \lk{(j, k)} is not evaluated, and we don't compute the exact demand
on the input map. For these specifications, we write a function that
takes an \lk{Evaluation} value, which contains the inputs, the
observed demand on the inputs, and a randomly generated demand on the
output by the framework.

The specification then checks whether the expected property holds for
this evaluation, and if the evaluation fails the expectation, it
outputs an evidence of the failure. The evidence is then returned to
the programmer by the framework for further investigation. In this
case, the evidence is simply the \lk{(j, k)} pair at which the
solution table was accessed.

\section{Demand Types and Specification}
\label{sec:demand-types-and-spec}
There is a general pattern between a type \lk{a} and
its \textit{demand type}, for which we will write \lk{Demand a} from
now on. In both the example on lists and the example on maps,
a \lk{Thunk} type constructor is placed around all of the constructor
arguments to the original type. The \lk{Thunk} type has the following
definition:
\begin{inlinecode}
data Thunk a = E !a | T
\end{inlinecode}
Its two constructors \lk{E} and \lk{T} represents an evaluated
fragment of the original data, and unexplored regions of the original
data, respectively.
%
This design is motivated by the runtime representation of Haskell
values. Boxed values in Haskell are indirectly represented by a
pointer, and the pointer can lead to either a constructor bundled with
pointers to deeper parts of the value, or a piece of unevaluated code,
which will produce a fragment of the value upon execution
~\cite{SPJSTG}. This piece of code is also know as a thunk.

@@

An \textit{exact specification} in StrictCheck is a function that
computes demand values. When we talk about Haskell function's laziness
informally, we often refer to these functions as ``strict'' or
``lazy'' in its arguments. There is a lot of information we can unpack
from such phrases --- how strict the function is in its arguments, the
dependency of the strictness on the argument values, and the
relationship between demand on inputs with demand on the output.

@@

In general, specification types are not unique for the same
function. Again consider \lk{take}, we can certainly choose the
specification type \lk{Spec '[Int] ([a] -> [a])} as well. However, the
meaning of this specification is very different from the other
choice. A specification of this type predicts how the \lk{Int}
argument is used given a demand on the partially-applied \lk{take}
function, and this specification does not produce information on how
the lists are used at all. StrictCheck specifications apply to
higher-order functions as well. Consider \lk{map :: (a -> b) -> [a] ->
[b]}, the type of its specification is \lk{Spec '[a -> b, [a]]
[b]}. In order to test higher-order functions, StrictCheck needs to
generate functions with interesting laziness behavior. This is
discussed in more details in \Cref{sec:gen-lazy-fun}.

The \lk{Spec} type has a single constructor with the same name. This
constructor takes a higher-order function in
continuation-passing-style ~\cn, where the first argument to the
function is a framework-supplied function that takes the predicted
demand on the inputs. We will always name this argument \lk{predict}
in this paper. The function is CPS'd to hide some internal plumbing
that presents a variadic interface to the user, so that the number of
arguments to \lk{predict} always matches the length of the type-level
list in \lk{Spec}.

@@

The \lk{Evaluation} type constructor has the same type-level index
as \lk{Spec}. Its first argument is the type-level list of all
argument types of the specified function, and the second argument is
the type of the specified function's result. Specifications
using \lk{Evaluation} must output values of type \lk{Maybe e},
where \lk{e} is the type of the evidence. StrictCheck is polymorphic
in this evidence type, so programmers can choose to output any useful
information for debugging. These specifications are spiritually
similar to properties in QuickCheck
~\cite{Claessen:2000:QLT:357766.351266}.
